\section{Sensitivity-Analysis Methodology}
\label{sec:sensitivity_methodology}
\jb{Reduce Sobol indices introduction, provide basic reserach of the 
sensitivity analysis methods and works dealing with sensitivity analysis 
involving random fracture netwroks.}

\subsection{Variance-based sensitivity analysis}
Our quantity of interest (QoI) $Y$ is the safety indicator produced
deterministically by the transport model. Hence, $Y=f(\mathbf{X})$ is a
function of the uncertain model parameters $\mathbf{X}=(X_1,\dots,X_D)$.
Assuming independence of the input parameters, the total variance of $Y$ can be
decomposed into contributions caused by individual parameters and by their
interactions.

Importantly, variance-based sensitivity analysis does \emph{not} require any
smoothness or continuity of $f$. This allows us to use a single parameter as a
seed for a pseudo-random generator and, in this way, generate
a (high-dimensional) vector of random inputs. We can then quantify the
aggregate impact of uncertainty coming from this parameter group. Ultimately,
such a vector may represent one realization of an entire random fracture
network or a random field, i.e.\ a~representation of aleatory uncertainty.

\paragraph{ANOVA variance decomposition.}
Under the independence assumption, the Hoeffding--Sobol (functional ANOVA)
decomposition holds in $L^2$:
\begin{equation}
f(\mathbf{X}) = f_0 + \sum_{i=1}^D f_i(X_i) + \sum_{1\le i<j\le D} f_{ij}(X_i,X_j) + \cdots + f_{1\cdots D}(X_1,\dots,X_D),
\end{equation}
with $f_0=\mathbb{E}[Y]$ and orthogonality conditions ensuring uniqueness, e.g.
$\mathbb{E}[f_u(\mathbf{X}_u)\,|\,X_k]=0$ for any $k\in u$. Consequently, the
variance decomposes as
\begin{equation}
\Var(Y) = \sum_{i=1}^D V_i + \sum_{1\le i<j\le D} V_{ij} + \cdots + V_{1\cdots D},
\end{equation}
where $V_u=\Var\!\big(f_u(\mathbf{X}_u)\big)$ is the partial variance
contributed by the subset $u\subseteq\{1,\dots,D\}$.

\paragraph{Sobol sensitivity indices: definitions and interpretation.}
The \emph{first-order} (main-effect) Sobol index of $X_i$ measures the fraction
of output variance explained by varying $X_i$ alone:
\begin{equation}
S_i = \frac{V_i}{\Var(Y)}.
\end{equation}
The \emph{second-order} index quantifies the additional variance due to
interaction between $X_i$ and $X_j$ beyond their main effects:
\begin{equation}
S_{ij} = \frac{V_{ij}}{\Var(Y)}.
\end{equation}
The \emph{total-effect} index measures the overall contribution of $X_i$,
including all its interactions with other inputs:
\begin{equation}
S_{T_i} = \frac{\sum_{u\ni i} V_u}{\Var(Y)},
\end{equation}
The exact indices satisfy inequalities:
$0\le S_i\le S_{T_i}\le 1$, and
\begin{equation}
\sum_{i=1}^D S_i \le 1.
\end{equation}
In particular, $S_{T_i}-S_i$ is a useful diagnostic of how
strongly $X_i$ interacts with other parameters. If $S_{T_i}$ is small, then
$X_i$ can often be fixed (or simplified) with minimal impact on $\Var(Y)$.
Higher-order indices, while theoretically possible, are usually not considered
in practice because they are prohibitively computationally demanding.

\paragraph{Estimation of Sobol indices and computational cost.}
Sobol indices are typically estimated by Monte Carlo sampling. A common choice
is the Saltelli sampling scheme \cite{Saltelli_Variance_2010} based on two
independent sampling matrices $A,B\in\mathbb{R}^{N\times D}$ and $D$
\textquote{mixed} matrices $A_B^{(i)}$ obtained by replacing the
$i$-th column of $A$ with the $i$-th column of $B$. This leads to a structured
set of model evaluations of size
\begin{equation}
N \times (2D+2).
\end{equation}
In practice, one often takes $N\gtrsim 10^3$ to obtain sufficient precision for
the dominant indices, whereas substantially larger $N$ may be needed to
estimate second-order indices reliably.

In our computations, a \emph{single} full model evaluation takes about
$30$ minutes. For $D=6$ parameters, the full Saltelli design therefore requires
tens of thousands of model evaluations, which in our case translates into
approximately one day on $400$ CPUs. This cost makes a naive extension to many
more parameters infeasible.

\paragraph{Towards reduced computational complexity.}
To make the approach tractable, we group inputs into a small number of
physically meaningful parameter groups and compare the variance
contributions \emph{between groups} (grouped Sobol indices), rather than
attempting to resolve all individual parameters at once.

Unfortunately the standard surrogate-model acceleration techniques
(e.g.\ polynomial chaos expansions) are not directly applicable
due to highly non-smooth and convoluted nature of the model function $f(\vc X)$
due to aleatory uncertainty encoded via pseudo-random seeds.

To reduce the cost of sensitivity analysis by at least two orders of magnitude,
we are developing variance-reduction and multi-fidelity strategies, in
particular numerical homogenization ideas \cite{Spetlik_Convolutional_2026} and
multi-level Monte Carlo (MLMC) estimators \cite{Giles2015}.

\subsection{Stochastic model parameters}
% The following summary is based on the data from technical reports of SÚRAO from several research projects studying \URFBII.

\subsubsection{Types of parameter distributions}
To describe the distributions of the independent model parameters $X_i$, we
specify a probability distribution for each parameter. Where applicable, each
distribution is characterized by
(i) a \emph{location} parameter (e.g. a mean) that sets the typical magnitude
of sampled values, and
(ii) an \emph{uncertainty scale} parameter (e.g., a standard deviation) that
quantifies sample variability.
We use following four distribution types.

\paragraph{Lognormal distribution}
\jb{Introduction of GM and CIF is nonstandard and must be carefully explained.}
Many transport-model parameters (e.g., hydraulic conductivity, diffusivity,
\dots) are assumed to be lognormally distributed. For lognormal variables, the
arithmetic mean and standard deviation are often not the most informative
descriptors. Following previous work \cite{tz_design}, we therefore use the
geometric mean \Gls{GM} as the location parameter, which has the same physical
unit as the quantity $Q$:
\[
\gm(Q) = \exp\!\Big(\mathrm{mean}\big(\log(Q)\big)\Big).
\]
As an uncertainty descriptor, we use the confidence-interval factor \Gls{CIF}:
\[
\cif(Q) = \exp\!\Big(3 \cdot \std\big(\log(Q)\big)\Big).
\]
The \Gls{CIF} admits a convenient interpretation: the (approximate) $99~\%$
confidence interval of $Q$ is
\[
\conf(Q,99\%) \approx \gm(Q)\,\Big[\cif^{-1}(Q),\, \cif(Q)\Big].
\]

\paragraph{Truncated normal distribution}
Porosity is modeled using a truncated normal distribution with mean as the
location parameter,  a truncation interval $(q_{\min}, q_{\max})$, and standard
deviation as the scale parameter. All distribution parameters have the same
unit as the modeled quantity.

\paragraph{Bayesian posterior distribution}
The parameters $(p_{30}, m, \alpha)$ of the DFN fracture families are not
independent and are therefore sampled jointly. We draw these parameters as
tuples from the empirical population of posterior samples obtained by Bayesian
inversion. Hence, the distribution is represented by this finite set of
samples.

\paragraph{Seed}
Aleatory uncertainties (fracture-network realization and computational mesh
generation) are represented via the seed of a pseudorandom number generator
driving the corresponding random processes. The seed is obtained by applying a
hash function to a random value drawn from the uniform distribution on $(0,1)$.
The hash is applied to decorrelate the seed from the raw sample value produced
by the same pseudorandom generator, ensuring reproducible yet effectively
independent realizations.

%JB: continue here
